---
title: "Mycobacterium Tuberculosis Project"
author: "Zaynab Mousavian"
date: "Sep. 2021"
output:
  pdf_document: default
  html_document:
    theme:
      bg: white
      fg: black
      base_font:
        google: Prompt
subtitle: Differential Expression Analysis
---
  
**Outline:**  
- Loading all data and packages
- Differential expression analysis  
- Showing Volcano plots (Figure 3)

---
# 1. Loading all data and packages   
After changing the current working directory, you can load all necessary packages and also the data, prepared in the Preprocessing.Rmd file, which is ready for any further analysis.

```{r, echo = TRUE,include=TRUE,message = FALSE,warning = FALSE}
knitr::opts_chunk$set(echo = TRUE,include=TRUE,message = FALSE,warning = FALSE)
percentage=0.3
inputDirectory <- paste0("../Results/Preprocessing/",as.character(percentage),"-")
resultsDirectory <- paste0("../Results/DifferentialExpressionAnalysis/",as.character(percentage),"-")

library(limma)
library(ggrepel)
library(ggplot2)
library(ggvenn)
library(reshape)
library(reshape2)
library(cowplot)
lnames <- load(paste0(inputDirectory,"ReadyData.RData"))
lnames
```


# 2. Differential expression analysis 
We used the Limma R package to identify differential expressed proteins between three groups of samples including: Active, Healthy controls and Latent.

```{r}
f <- factor(groups)
design <- model.matrix(~0+f)
colnames(design) <- c("Active", "Latent", "Control")
cont.matrix <- makeContrasts(Active-Control,Latent-Control,Active-Latent,levels=design)
fit <- lmFit(dat_Expr, design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit.b <- eBayes(fit2)
save(fit.b,file=paste0(resultsDirectory,"DE-Results.RData"))
```

# 3. Showing Volcano plots (Figure 3)

```{r}
volcanoPlot_func <- function(inputData, xaxis, yaxis,xlab) {
  inputData$sig <- ifelse(inputData[,yaxis] <= 0.05&abs(inputData[,xaxis]) > 1, inputData$sig <- "Sig", inputData$sig <- "nSig")
  ggplot(inputData, aes(x=.data[[xaxis]],y=-log10(.data[[yaxis]])))+
    geom_point(aes(color = sig,size=2)) +
    scale_color_manual(values = c("grey", "red")) +
    theme_bw(base_size = 12) +
    geom_text_repel(
      data = subset(inputData, inputData[,yaxis] <= 0.05 & abs(inputData[,xaxis]) > 1), 
      aes(label = Row.names),
      size = 10,
      box.padding = unit(0.35, "lines"),
      point.padding = unit(0.3, "lines"),
      segment.colour = "transparent") +
    labs(y="-Log10(FDR)", x=paste0("Log2(",xlab,")"))+
  theme(legend.position = "none",axis.title = element_text(size=22,face="bold"),axis.text = element_text(size=18)) 
}

logFC <- as.data.frame(fit.b$coefficients)
pValue <- as.data.frame(fit.b$p.value)
pValue2 <- apply(pValue,2,function(x) p.adjust(x,method = "fdr"))
volcanoPlot_data <- cbind(logFC=logFC,pValue=pValue2)
names(volcanoPlot_data) <- lapply(names(volcanoPlot_data), function(x) gsub(" - ",".",x))
volcanoPlot_data$sig <- ifelse(volcanoPlot_data$pValue.Active.Control < 0.05&abs(volcanoPlot_data$logFC.Active.Control)>1, volcanoPlot_data$sig <- "Sig", volcanoPlot_data$sig <- "nSig")
volcanoPlot_data$Row.names <- row.names(volcanoPlot_data)
volcanoPlot1 <- volcanoPlot_func(volcanoPlot_data,"logFC.Active.Control","pValue.Active.Control","Active vs. Control")
volcanoPlot2 <- volcanoPlot_func(volcanoPlot_data,"logFC.Active.Latent","pValue.Active.Latent","Active vs. Latent")
volcanoPlot3 <- volcanoPlot_func(volcanoPlot_data,"logFC.Latent.Control","pValue.Latent.Control","Latent vs. Control")

pdf(paste0(resultsDirectory,"Figure 3.pdf"),height = 10,width = 20)
cowplot::plot_grid(volcanoPlot1,volcanoPlot2,ncol = 2,nrow = 1)
dev.off()
```

