---
title: "Mycobacterium Tuberculosis Project"
author: "Zaynab Mousavian"
date: "Oct. 2021"
output:
  pdf_document: default
  html_document:
    theme:
      bg: white
      fg: black
      base_font:
        google: Prompt
subtitle: SignatureValidation
---
  
**Outline:**  
- Loading data from the curatedTBData package  
- Loading all data and R packages
- Comparing all modules using GSVA package
- Showing results of all methods on all modules (Figures S8-S10) 
- Validating our signature using ssGSEA method
- Showing ssGSEA results (Figure 5A)
- Running wilcox test on ssGSEA results 
- Validating our signature using QuSAGE method
- Showing all QuSAGE results (Figure 5B)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Loading data from the curatedTBData package 
we used the transcriptomic TB datasets from the curatedTBData R package to validate our signature in this study. We loaded all datat provided in this package and saved it as an R object, then used in further steps.

```{r}
library(curatedTBData)
library(MultiAssayExperiment)
data(DataSummary)
row.names(DataSummary)=DataSummary$Study
object_list <- curatedTBData(DataSummary$Study, dry.run = FALSE, curated.only = FALSE)
save(object_list, DataSummary, file="../Datasets/curatedTBData/curatedTBData.RData")
```

# 2. Loading all data and R packages
At first, we load all R packages, our 12-protein signature and all TB transcriptomics datasets for validating the signature.

```{r}
library(GSVA)
library(CePa)
library(qusage)
library(ggplot2)
library(limma)
library(cowplot)
library(reshape2)
library(ggrepel)

percentage <- 0.3
corMeasure <- "pearson"
run <- TRUE
inputDirectory <- paste0("../Results/Coexpression/",
                           as.character(percentage),"-",corMeasure,"-")
load(paste0(inputDirectory,"modules.Rdata"))


modules[[4]]=gsub("-","",modules[[4]])
modules[[4]][3]="IFNG"
modules[[4]][9]="CCL7"
modules[[4]][14]="TGFB1"
modules[[4]][15]="CD274"
modules[[4]]=modules[[4]][-c(6,8,12,14)]

modules[[1]][4]="EIF4EBP1"
modules[[1]][3]="SULT1A1"

modules[[2]][4]="S100A12"
modules[[2]][5]="TGFA"

load("../Datasets/curatedTBData/curatedTBData.RData")
```

# 3. Comparing all modules using GSVA package
We applied all four modules to several independent cohorts using multiple gene set enrichment analysis methods. Nine transcriptomic TB datasets were selected based on the criteria of age >15, number of individuals per group >10 and no anti-TB treatment.

```{r}
studies=c("GSE73408", "GSE101705", "GSE19439", "GSE19442", "GSE19444",  "GSE42826", "GSE42830", "GSE83456", "GSE28623")
GSE=data.frame()
gsea = function(studyName){

  data_Expr = assay(object_list[[studyName]])
  boxplot(data_Expr)
  pheno_Data = colData(object_list[[studyName]])
  data_Traits= as.data.frame(pheno_Data[,"TBStatus",drop=FALSE])
  names(data_Traits)="Type"
  
  data_Traits$Type=gsub("Control","Control",data_Traits$Type)
  data_Traits$Type=gsub("LTBI","Latent",data_Traits$Type)
  data_Traits$Type=gsub("PTB","Active",data_Traits$Type)
  data_Traits$Type=gsub("OD","Other",data_Traits$Type)

  GSE_func = function(methodName){
    GSE_method <- methodName
    GSE_results = gsva(as.matrix(data_Expr),modules,method=GSE_method)
    row.names(GSE_results) = modNames
    modNames_New = modNames[!modNames=="grey"]
    GSE_results = GSE_results[modNames_New,]
    GSE_results_melted = melt(GSE_results)
    names(GSE_results_melted) = c("Module","Sample","Score")
    GSE_results_melted$Type = unlist(lapply(GSE_results_melted$Sample, function(x) data_Traits[x,"Type"]))
    GSE_results_melted$Method = GSE_method
    GSE_results_melted$Dataset = studyName
    GSE <<- rbind(GSE,GSE_results_melted)
  }
  lapply(c("ssgsea","gsva","zscore"),function(x) GSE_func(x))
}
lapply(studies, function(x) gsea(x))

GSE=GSE[-which(GSE$Type=="Other"),]
GSE=GSE[-which(GSE$Type=="Subclinical"),]
GSE=GSE[-which(is.na(GSE$Type)),]
GSE$Type = gsub("Healthy","Control",GSE$Type)
GSE$Type <- factor(GSE$Type, levels = c("Active", "Latent", "Control"))
GSE$Method <- factor(GSE$Method, levels = c("ssgsea", "gsva", "zscore"))
GSE$Module <- factor(GSE$Module, levels = c("turquoise", "blue", "brown","yellow"))
save(GSE,file="../Results/GeneSetEnrichment/curatedTBData-ALLModules-ALLMethods.RData")
```

# 4. Showing results of all methods on all modules (Figures S8-S10)
The results of three enrichment methods (ssgsea, gsva and zscore) on all modules were shown in this section.

```{r}
load("../Results/GeneSetEnrichment/curatedTBData-ALLModules-ALLMethods.RData")
method="ssgsea"
GSE2=GSE[which(GSE$Method==method),]
pdf(file=paste0("../Results/GeneSetEnrichment/Figure S8.pdf"),width=35,height = 30)

ggplot(GSE2,aes(x=Module,y=Score,fill=Type))+
  stat_boxplot(geom ='errorbar') +
        geom_boxplot()+
      scale_y_continuous(limits = c(-1,1))+
        facet_wrap(~Dataset,scales ="fixed",nrow = 3,ncol=3)+
        theme(axis.title.x = element_blank(), axis.title = element_text(size=40,face="bold"),axis.text=element_text(size=30),strip.text = element_text(size=40, face=2),legend.key.size = unit(4, 'cm'),legend.text = element_text(size=30),legend.title = element_blank())+
        scale_fill_manual(values = c("#2F8AC4", "#E48725", "#A5AA99"))+
  labs(y=paste0("Enrichment score using ",method," method"))

dev.off()

method="gsva"
GSE2=GSE[which(GSE$Method==method),]
pdf(file=paste0("../Results/GeneSetEnrichment/Figure S9.pdf"),width=35,height = 30)

ggplot(GSE2,aes(x=Module,y=Score,fill=Type))+
  stat_boxplot(geom ='errorbar') +
        geom_boxplot()+
      scale_y_continuous(limits = c(-1,1))+
        facet_wrap(~Dataset,scales ="fixed",nrow = 3,ncol=3)+
        theme(axis.title.x = element_blank(), axis.title = element_text(size=40,face="bold"),axis.text=element_text(size=30),strip.text = element_text(size=40, face=2),legend.key.size = unit(4, 'cm'),legend.text = element_text(size=30),legend.title = element_blank())+
        scale_fill_manual(values = c("#2F8AC4", "#E48725", "#A5AA99"))+
  labs(y=paste0("Enrichment score using ",method," method"))

dev.off()

method="zcore"
GSE2=GSE[which(GSE$Method==method),]
pdf(file=paste0("../Results/GeneSetEnrichment/Figure S10.pdf"),width=35,height = 30)

ggplot(GSE2,aes(x=Module,y=Score,fill=Type))+
  stat_boxplot(geom ='errorbar') +
        geom_boxplot()+
      scale_y_continuous(limits = c(-1,1))+
        facet_wrap(~Dataset,scales ="fixed",nrow = 3,ncol=3)+
        theme(axis.title.x = element_blank(), axis.title = element_text(size=40,face="bold"),axis.text=element_text(size=30),strip.text = element_text(size=40, face=2),legend.key.size = unit(4, 'cm'),legend.text = element_text(size=30),legend.title = element_blank())+
        scale_fill_manual(values = c("#2F8AC4", "#E48725", "#A5AA99"))+
  labs(y=paste0("Enrichment score using ",method," method"))

dev.off()
```


# 5. Validating our signature using ssGSEA method
To assess the significance of the 12-marker signature in independent TB cohorts, we performed ssGSEA on nine transcriptomic TB datasets from the curatedTBData R package.

```{r}
modules[[4]]=modules[[4]][-c(6,8,12,14)]
studies=c("GSE73408", "GSE101705", "GSE19439", "GSE19442", "GSE19444",  "GSE42826", "GSE42830", "GSE83456", "GSE28623")

f=function(studyName){
  data_Expr = assay(object_list[[studyName]])
  dim(data_Expr)
  modules[[4]][which(modules[[4]]%in%row.names(data_Expr))]
  length(which(modules[[4]]%in%row.names(data_Expr)))
}
lapply(studies, function(x) f(x))


GSE=data.frame()
gsea = function(studyName){
  data_Expr = assay(object_list[[studyName]])
  pheno_Data = colData(object_list[[studyName]])
  data_Traits= as.data.frame(pheno_Data[,"TBStatus",drop=FALSE])
  names(data_Traits)="Type"

  data_Traits$Type=gsub("Control","Control",data_Traits$Type)
  data_Traits$Type=gsub("LTBI","Latent",data_Traits$Type)
  data_Traits$Type=gsub("PTB","Active",data_Traits$Type)
  data_Traits$Type=gsub("OD","Other",data_Traits$Type)
  
  length(which(modules[[4]]%in%row.names(data_Expr)))
  GSE_func = function(methodName){
    GSE_method <- methodName
    GSE_results = gsva(as.matrix(data_Expr),modules,method=GSE_method)
    row.names(GSE_results) = modNames
    modNames_New = modNames[modNames=="turquoise"]
    GSE_results = GSE_results[modNames_New,,drop=FALSE]
    GSE_results_melted = melt(GSE_results)
    names(GSE_results_melted) = c("Module","Sample","Score")
    GSE_results_melted$Type = unlist(lapply(GSE_results_melted$Sample, function(x) data_Traits[x,"Type"]))
    GSE_results_melted$Method = GSE_method
    GSE_results_melted$Dataset = studyName
    GSE <<- rbind(GSE,GSE_results_melted)
  }
  lapply(c("ssgsea"),function(x) GSE_func(x))
}
lapply(studies, function(x) gsea(x))
save(GSE,file=paste0(outputDirectory,"ssGSEA-curatedTBData.RData"))
```

# 6. Showing ssGSEA results (Figure 5A)

```{r}
lnames=load(paste0(outputDirectory,"ssGSEA-curatedTBData.RData"))
GSE=GSE[-which(GSE$Type=="Other"),]
#GSE=GSE[-which(GSE$Type=="Subclinical"),]
GSE=GSE[-which(is.na(GSE$Type)),]
GSE$Type = gsub("Healthy","Control",GSE$Type)
GSE$Type = gsub("EActive","Active",GSE$Type)
GSE$Type <- factor(GSE$Type, levels = c("Active", "Latent", "Control"))

pdf(file=paste0(outputDirectory,"Figure 5A.pdf"),width=15,height = 20)

ggplot(GSE,aes(x=Type,y=Score,fill=Type))+
             geom_violin() + geom_jitter(height = 0, width = 0.1)+
                scale_y_continuous(limits=c(-0.4,0.4)) + 
                facet_wrap(~Dataset,scales = "fixed",nrow = 3,ncol = 3)+ 
               theme(axis.title.x = element_blank(), axis.title = element_text(size=30,face="bold"),axis.text =element_text(size=20),strip.text = element_text(size=30, face=2))+
        scale_fill_manual(values = c("#2F8AC4", "#E48725", "#A5AA99"))+
  labs(y=paste0("Enrichment score using ssGSEA method"))
 
dev.off()
```

# 7. Running wilcox test on ssGSEA results 
We used the wilcox test to assess the statistical significance for the enrichment score difference between either active TB and Latent TB or Active TB and healthy controls.

```{r}
data=data.frame()
diff=function(studyName){
  selected=GSE[which(GSE$Dataset==studyName),]
  score=selected$Score
  f <- factor(selected$Type)
  res1<- t.test(score[which(selected$Type=="Active")],score[which(selected$Type=="Control")])
  res2<- wilcox.test(score[which(selected$Type=="Active")],score[which(selected$Type=="Control")])
  data <<- rbind(data,cbind(pVal1=res1$p.value,pVal2=res2$p.value,name=studyName))
}
lapply(studies[c(3,5:9)],function(x) diff(x))
lapply(studies[c(1:5,9)],function(x) diff(x))
```

# 8. Validating our signature using QuSAGE method
Moreover, to verify that the protein signature was specific to active TB and not to other non-TB respiratory infections or diseases with clinical presentations similar to active TB, the qusage function from the QuSAGE R package was applied to datasets comparing active TB or non-TB disease to various control groups.

## 8.1. Using TB transcriptomics datasets

```{r}
studies=c("GSE73408", "GSE101705", "GSE19439", "GSE19442", "GSE19444",  "GSE42826", "GSE42830", "GSE83456", "GSE28623")

GSE=data.frame()
gsea = function(studyName){
  data_Expr = assay(object_list[[studyName]])
  pheno_Data = colData(object_list[[studyName]])
  data_Traits= as.data.frame(pheno_Data[,"TBStatus",drop=FALSE])
  names(data_Traits)="Type"
  data_Traits$Type=gsub("Control","Control",data_Traits$Type)
  data_Traits$Type=gsub("LTBI","Latent",data_Traits$Type)
  data_Traits$Type=gsub("PTB","Active",data_Traits$Type)
  data_Traits$Type=gsub("OD","Other",data_Traits$Type)
  
  QuSAGE_func = function(contrastName){
    QuSAGE_results = qusage(as.matrix(data_Expr),data_Traits$Type, contrastName,modules)
    res = qsTable(QuSAGE_results)
    plotCIsGenes(QuSAGE_results,path.index = 4,main = paste0(studyName,":",contrastName))
    res=res["4",]
    GSE <<- rbind(GSE,cbind(studyName,Comparison=contrastName,logFC = res$log.fold.change , pVal=res$p.Value, FDR=res$FDR))
  }
  if (length(which(data_Traits$Type=="Active"))!=0){
    if (length(which(data_Traits$Type=="Control"))!=0){
      QuSAGE_func("Active-Control")
    }
    else{
      GSE <<- rbind(GSE,cbind(studyName,Comparison="Active-Control",logFC = NA , pVal=NA, FDR=NA))
    }
    if (length(which(data_Traits$Type=="Latent"))!=0){
      QuSAGE_func("Active-Latent")
    }
    else{
      GSE <<- rbind(GSE,cbind(studyName,Comparison="Active-Latent",logFC = NA , pVal=NA, FDR=NA))
    }
  }
}
lapply(studies, function(x) gsea(x))
GSE$logFC=as.numeric(GSE$logFC)
GSE$pVal=as.numeric(GSE$pVal)
GSE$FDR = as.numeric(GSE$FDR)
save(GSE,file = paste0(outputDirectory,"QuSAGE-curatedTBData.RData"))
dev.off()
```

## 8.2 Using LRTI datasets 

```{r}
studies = c("GSE42026","GSE40012","GSE60244")

f=function(studyName){
  load(file=paste0("../Datasets/",studyName,"/ReadyData.RData"))
  modules[[4]][which(modules[[4]]%in%row.names(data_Expr))]
}
lapply(studies, function(x) f(x))

GSE= data.frame()
gsea = function(studyName){
 
  if (studyName == "GSE40012"){
    l=load(file=paste0("../Datasets/",studyName,"/ReadyData-Day0.RData"))
  }
  else{
    l=load(file=paste0("../Datasets/",studyName,"/ReadyData.RData"))
  }
  QuSAGE_func = function(contrastName){
    
    QuSAGE_results = qusage(as.matrix(data_Expr),data_Traits$Type, contrastName,modules)
    res = qsTable(QuSAGE_results)
    res=res["4",]
    GSE <<- rbind(GSE,cbind(studyName,Comparison=contrastName,logFC = res$log.fold.change , pVal=res$p.Value, FDR=res$FDR))
  }
  if (studyName == "GSE60244"){
    lapply(c("Virus-Control","Coinfection-Control","Bacteria-Control"), function(x) QuSAGE_func(x))
  }
  if (studyName == "GSE40012"){
    lapply(c("Bacterial-Control","InfluenzaA-Control","SIR-Control"), function(x) QuSAGE_func(x))
  }
  if (studyName == "GSE42026"){
    lapply(c("Bacterial-Control","H1N109-Control","RSV-Control"), function(x) QuSAGE_func(x))
  }
}
lapply(studies, function(x) gsea(x))
GSE$logFC=as.numeric(GSE$logFC)
GSE$pVal=as.numeric(GSE$pVal)
GSE$FDR = as.numeric(GSE$FDR)
save(GSE,file = paste0(outputDirectory,"QuSAGE-LRTI-datasets.RData"))
```

## 8.3. Using Sarcoidosis datasets 

```{r}
studies=c("GSE83456","GSE42830","GSE42826")
f=function(studyName){
  data_Expr = assay(object_list[[studyName]])
  modules[[4]][which(modules[[4]]%in%row.names(data_Expr))]
 
}
lapply(studies, function(x) f(x))

GSE= data.frame()
gsea = function(studyName){
 
  data_Expr = assay(object_list[[studyName]])
  pheno_Data = colData(object_list[[studyName]])
  data_Traits= as.data.frame(pheno_Data[,"TBStatus",drop=FALSE])
  names(data_Traits)="Type"
  
  data_Traits$Type=gsub("Control","Control",data_Traits$Type)
  data_Traits$Type=gsub("LTBI","Latent",data_Traits$Type)
  data_Traits$Type=gsub("PTB","Active",data_Traits$Type)
  data_Traits$Type=gsub("OD","Other",data_Traits$Type)
  
  control_Samples = row.names(data_Traits[data_Traits$Type=="Control",,drop=FALSE])
  disease_Samples = row.names(data_Traits[data_Traits$Type=="Other",,drop=FALSE])
  data_Expr=data_Expr[,c(control_Samples,disease_Samples)]
  data_Traits=data_Traits[c(control_Samples,disease_Samples),,drop=FALSE]
  
  QuSAGE_results = qusage(as.matrix(data_Expr),data_Traits$Type, "Other-Control",modules)
  res = qsTable(QuSAGE_results)
  res=res["4",]
  GSE <<- rbind(GSE,cbind(studyName,Comparison="Other-Control",logFC = res$log.fold.change , pVal=res$p.Value, FDR=res$FDR))
  
}
lapply(studies, function(x) gsea(x))
GSE$logFC=as.numeric(GSE$logFC)
GSE$pVal=as.numeric(GSE$pVal)
GSE$FDR = as.numeric(GSE$FDR)
save(GSE,file = paste0(outputDirectory,"QuSAGE-Sarcoidosis-datasets.RData"))
```

# 9. Showing all QuSAGE results (Figure 5B)

```{r}
load(paste0(outputDirectory,"QuSAGE-Sarcoidosis-datasets.RData"))
GSE_0=GSE
GSE_0$Type_of_Infection="Other"
load(paste0(outputDirectory,"QuSAGE-LRTI-datasets.RData"))
GSE_1=GSE
GSE_1$Type_of_Infection="Other"
load(paste0(outputDirectory,"QuSAGE-curatedTBData.RData"))
GSE$Type_of_Infection="Tuberculosis"
GSE_1$Type_of_Comparison = "other infection vs. controls"
GSE_1[which(GSE_1$Comparison=="Bacterial-Control"),"trick"]=":1"
GSE_1[which(GSE_1$Comparison=="H1N109-Control"),"trick"]=":2"
GSE_1[which(GSE_1$Comparison=="RSV-Control"),"trick"]=":3"
GSE_1[which(GSE_1$Comparison=="Bacterial-Control"),"trick"]=":1"
GSE_1[which(GSE_1$Comparison=="InfluenzaA-Control"),"trick"]=":2"
GSE_1[which(GSE_1$Comparison=="SIR-Control"),"trick"]=":3"
GSE_1[which(GSE_1$Comparison=="Virus-Control"),"trick"]=":1"
GSE_1[which(GSE_1$Comparison=="Coinfection-Control"),"trick"]=":2"
GSE_1[which(GSE_1$Comparison=="Bacteria-Control"),"trick"]=":3"
GSE[which(GSE$Comparison=="Active-Control"),"Type_of_Comparison"]="active TB vs. controls"
GSE[which(GSE$Comparison=="Active-Control"),"trick"]=":1"
GSE[which(GSE$Comparison=="Active-Latent"),"Type_of_Comparison"]="active TB vs. latent TB"
GSE[which(GSE$Comparison=="Active-Latent"),"trick"]=":2"
GSE_0$Type_of_Comparison = "other disease vs. controls"
GSE_0[which(GSE_0$Comparison=="Other-Control"),"trick"]=":2"
GSE=rbind(GSE,GSE_1,GSE_0)
GSE=GSE[-which(is.na(GSE$logFC)),]
GSE$Value = -log10(GSE$FDR)
sep = ifelse(unlist(lapply(GSE$studyName,function(x) nchar(x)))==8,"  ","")
GSE$Name = paste0 (GSE$studyName,sep,GSE$trick)
GSE = GSE[order(GSE[,"Value"],decreasing = FALSE),,drop=FALSE]
GSE$Name <- factor(GSE$Name, levels = GSE$Name[order(GSE[,"Value"],decreasing = FALSE)])

pdf(file=paste0(outputDirectory,"Figure 5B.pdf"),width = 8,height = 10)
ggplot(GSE,aes(x = Value, y = Name,group=1))+
  geom_point(aes(shape = Type_of_Comparison,size=6)) +
  scale_shape_manual(values = c(0,1,3,4)) +
  coord_cartesian(xlim = c(0, 15))+
  geom_vline(xintercept = 1.30 , color = "red", linetype = "dashed")+
  labs( x=paste0("-Log10(FDR) using QuSAGE method"))+
  theme(axis.title = element_text(size=14,face="bold"),axis.text = element_text(size=12))
dev.off()
```

