---
title: "Mycobacterium Tuberculosis Project"
author: "Zaynab Mousavian"
date: "Oct. 2021"
output:
  pdf_document: default
  html_document:
    theme:
      bg: white
      fg: black
      base_font:
        google: Prompt
subtitle: SignatureOverlap
---
  
**Outline:**  
- Loading all data and R packages 
- Comparing our signature with published TB signatures (Figure 5C)
- Showing overlap between published TB signatures (Figure S7)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Loading all data and R packages 
We firstly load all TB signatures, from the TBSignatureProfiler R package, and necessary R packages.

```{r}
suppressPackageStartupMessages({
  library(TBSignatureProfiler)
  library(SummarizedExperiment)
  library(curatedTBData)
})
library(ggplot2)
library(reshape)
load("../Results/Coexpression/0.3-pearson-modules.RData")
modules[[4]]=gsub("-","",modules[[4]])
modules[[4]][3]="IFNG"
modules[[4]][9]="CCL7"
modules[[4]][14]="TGFB1"
modules[[4]][15]="CD274"
data=data.frame()
modules[[4]]=modules[[4]][-c(6,8,12,14)]
```

# 2. Comparing our signature with published TB signatures (Figure 5C)
We compared the 12-marker signature with the other published gene signatures from the TBSignatureProfiler R package and published protein signatures to investigate overlap between protein and transcriptional signatures with the proteins of our signature.

```{r}
overlap = function(gene){
  data <<- rbind(data,unlist(lapply(TBsignatures, function(x) ifelse(length(grep(gene,x))!=0,1,0))))
}
lapply(modules[[4]],function(x) overlap(x))
names(data)=names(TBsignatures)
row.names(data)=modules[[4]]
file=read.csv("../Manuscripts/Gene signatures/Summary.csv",sep=";")
studies = file[which(file$Include=="Yes"),"Study"]
data=data[,studies]

file=read.csv("../Manuscripts/Gene signatures/Protein/new/Summary.csv",sep=";")
file=file[file$Include.in.manuscript=="Yes",]
data2=data.frame(genes=modules[[4]])
lapply(file$Name,function(x) data2<<- cbind(data2,x))
row.names(data2)=data2$genes
data2=data2[,-1]
names(data2)=data2[1,]
data2[,]=0
data2["IL6","Ahmad_4"]=2
data2["IL6","Groote_6"]=2
data2["IL6","Pedersen_6"]=2
data2["VEGFA","Ahmad_4"]=2
data2["VEGFA","Delemare_4"]=2
data2["VEGFA","Pedersen_6"]=2
data2["IFNG","Chegou_7"]=2
data2["IFNG","Morris_9"]=2


data=cbind(data2,data)


colors=RColorBrewer::brewer.pal(8, "Dark2")
data$gene=row.names(data)
data=melt(data,id.vars = "gene")
data$value=factor(data$value)
pdf(file="../Results/Coexpression/Figure 5C.pdf",height = 5,width=15)
ggplot(data,aes(x=variable,y=gene,fill=value))+
  geom_tile(color="grey",size=0.15)+
  scale_fill_manual(values=c("white", colors[1],colors[3]))+
  theme(axis.text.x  = element_text(angle=90, hjust=1,vjust=0.5,size=10,face="bold"),axis.text.y.left  = element_text(hjust=0,size=10,face="bold"))
dev.off()
```

# 3. Showing overlap between published TB signatures (Figure S7)
We generate a heatmap plot to show the overlap in markers between studies investigating gene signatures aiming to discriminate active TB from latent TB, healthy controls, or other diseases.

```{r}
data=data.frame()

overlap = function(gene){
  data <<- rbind(data,unlist(lapply(TBsignatures, function(x) 100*length(intersect(gene,x))/min(length(gene),length(x)))))
}
lapply(TBsignatures,function(x) overlap(x))

names(data)=names(TBsignatures)
row.names(data)=names(TBsignatures)
data$gene=row.names(data)
data=melt(data,id.vars = "gene")
data$text=lapply(data$value, function(x) ifelse(x!=0,x,""))

pdf(file="../Results/Coexpression/Figure S7.pdf",height = 15,width=15)
ggplot(data,aes(x=variable,y=gene,fill=value))+
  geom_tile(color="grey",size=0.15)+
  scale_fill_gradient(low = "white", high = "#CD3A8E") +
  theme(axis.text.x  = element_text(angle=90, hjust=1,vjust=0.5,size=10,face="bold"),axis.text.y.left  = element_text(hjust=0,size=10,face="bold"))
dev.off()
```

